---
title: "ECOL 620 Lab 7"
author: "Natalie Schmer"
date: "3/18/2021"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r, include=F, echo=F, message=F}
{
  library(raster)          
  library(reshape2)         
  library(rgdal)            
  library(adehabitatLT)     
  library(adehabitatHR)     
  library(adehabitatHS)     
  library(survival)         
  library(tidyverse)
  library(scales)
}
```


# 1. 
After removing the coordinates, the panthers dataframe has the CatID, AgeClass, and Julian date. The CatID is a unique identifyer for each individual, the AgeClass (either adult or subadult), and the Julian Date is the number of days since some arbitrary date (possibly beginning of the study period), needed for the Brownian Bridge model. 
```{r, eval=F, include=F, echo=F}
panthers <- readOGR("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab7/panthers.shp")
panthers=panthers[,3:5]
panthers@data
```

# 2.
Lines 80-82 are using a 5 km moving window to determine the proportion of wet and dry woods within 5 km neighborhood, to figure out if the animal selecting the pixel or neighborhood around pixel for each of the forest types. `fw` is the function to set the weight around each point function that's using the 5 km radius, and `fw` is then used in the focal function for each forest type to calculate the portion of each forest type in the 5 km radius by lining up with the pixels and multiplying by weight from fw -- if everything within the 5 km buffer is the cover of interest then sum will be 1, otherwise will be some value between 0 and 1. 


# 3. MCP for panther 137: 


```{r, include=FALSE, message=F, warning=F, echo=FALSE}
## Data prep

#set working directory where data were downloaded
setwd("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab7")

###################################################
#8.3.2 Prepping the data
###################################################
#label projection for later use
crs.land <- "+proj=aea +lat_0=24 +lon_0=-84 +lat_1=24 +lat_2=31.5 +x_0=400000 +y_0=0 +ellps=GRS80 +datum=WGS84 +units=m +no_defs"

#landcover source: fwc/fnai
land <- raster("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab7/panther_landcover")

plot(land)

#check projection
projection(land)

#Add panther data
panthers <- readOGR("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab7/panthers.shp")
panthers=spTransform(panthers, crs.land)
plot(panthers)

#the x and y variables are likely in a different CRS, so I have removed them to avoid confusion
panthers=panthers[,3:5]

#check projection
projection(panthers)

#inspect
summary(panthers)
unique(panthers$CatID) #the unique cat IDs, Julian date for BBMM model 
head(panthers)

#plot
plot(land)
points(panthers, col=panthers$CatID)
unique(land)

#load reclassification table for reclassifying map
classification <- read.table("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab7/landcover reclass.txt", header=TRUE)

#inspect
head(classification)
classification$Description    #original classification
classification$Description2   #re-class

#format for reclassify function;
class <- as.matrix(classification[,c(1,3)])
land_sub <- reclassify(land,rcl=class)

#plot
plot(land_sub)

#create forested wetlands layer
wetforest <- land_sub
values(wetforest) <- 0
wetforest[land_sub==9 | land_sub==11] <- 1

#create forested uplands layer
dryforest <- land_sub
values(dryforest) <- 0
dryforest[land_sub==10 | land_sub==12] <- 1

#5 km moving window to get neighborhood proportion, is animal selecting pixel or neighborhood around pixel? good for niche modeling
fw <- focalWeight(land_sub, 5000, 'circle')
dry.focal <- focal(dryforest, w=fw, fun="sum", na.rm=T)
wet.focal <- focal(wetforest, w=fw, fun="sum", na.rm=T)

#merge into a single raster stack
layers <- stack(land_sub, wet.focal, dry.focal)
names(layers) <- c("landcover", "wetforest", "dryforest")

#plot
plot(layers)
```


```{r, include=FALSE, message=F, warning=F, echo=FALSE}
## Extract 50%, 75%, 95% MCPs for panther 137 
mcp95 <- mcp(panthers[,"CatID"], percent = 95)
mcp50 <- mcp(panthers[,"CatID"], percent = 50)
mcp75 <- mcp(panthers[,"CatID"], percent = 75)

mcp95_df <- fortify(mcp95, region='id')
mcp75_df <- fortify(mcp75, region='id')
mcp50_df <- fortify(mcp50, region='id')


#Selecting the cat
mcp95_df_137=subset(mcp95_df, id==137)
mcp75_df_137=subset(mcp75_df, id==137)
mcp50_df_137=subset(mcp50_df, id==137)

panther137_points <- as.data.frame(panthers) %>% 
                        filter(CatID == "137")
```


```{r, message=F, warning=F, echo=FALSE}
ggplot() +
  geom_polygon(data=mcp95_df_137,aes(x=long, y=lat),colour="black", fill="yellow")+
    geom_polygon(data=mcp75_df_137,aes(x=long, y=lat),colour="black", fill="green")+
  geom_polygon(data=mcp50_df_137,aes(x=long, y=lat),colour="black", fill="blue")+
  geom_point(panther137_points, mapping = aes(x = coords.x1, y= coords.x2))+
  ggthemes::theme_few()+
  labs(x = "Easting (m)",
       y = "Northing (m)",
       title = "Minimum Convex Hull Polygon: Panther 137")
```

# 4. Reclassified raster, 50% and 95% MCP for all cats 
```{r, include=FALSE, message=F, warning=F, echo=FALSE}
# reclassified needs to be points
land_sub_pt <- rasterToPoints(land_sub)
land_sub_df <- as.data.frame(land_sub_pt)
head(land_sub_df)

# need to join the reclass ids to the land_sub_df
land_sub_df <- classification %>% 
                  select(gfchab_500cl2 = ChangeTo, Description2) %>% 
                  left_join(land_sub_df, ., by = "gfchab_500cl2")
```

## Plot
```{r, message=F, warning=F, echo=FALSE}
  ggplot() +
    geom_raster(land_sub_df, mapping = aes(x, y, fill = Description2))+
    geom_polygon(data=mcp95_df,aes(x=long, y=lat, group = id), colour="black", fill = NA)+
    geom_polygon(data=mcp50_df,aes(x=long, y=lat, group = id), colour="black")+
  ggthemes::theme_few()+
  scale_x_continuous(label=scientific,breaks=c(600000,725000))+
scale_y_continuous(label=scientific, breaks=c(200000,300000,400000))+
  labs(fill = "Landcover Class",
       x = "Easting (m)",
       y = "Northing (m)")
```

# 5. All panthers 95% home ranges 

```{r, include=FALSE, message=F, warning=F, echo=FALSE}
## A. Min convex hull
# minimum convex hull
mch_panther_all_holes=subset(mcp95_df, hole==T)
mch_panther_all=subset(mcp95_df, hole==F)

(mch.plot <- ggplot()+
geom_polygon(data=mch_panther_all,aes(x=long, y=lat, group=group, fill=id), alpha=.8)+
geom_polygon(data=mch_panther_all_holes,aes(x=long, y=lat, group=group), fill="white", alpha=1)+
  ggthemes::theme_few()+
  scale_x_continuous(label=scientific,
                     breaks= c(600000, 725000),
                     limits = c(600000, 725000) )+
scale_y_continuous(label=scientific, 
                   breaks=c(200000,300000,400000),
                   limits = c(200000,400000))+
  labs(x = "Easting (m)",
       y = "Northing (m)",
       fill = "Panther ID",
       title = "Minimum Convex Hull Polygon"))
```


```{r, include=FALSE, message=F, warning=F, echo=FALSE}
## B. Bivariate Kernel

#kernel types: 
kernel.href.bivar <- kernelUD(panthers[,"CatID"], h="href", kern="bivnorm")

kernel.href.epa <- kernelUD(panthers[,"CatID"], h="href", kern="epa")

#plot
image(kernel.href.bivar)
image(kernel.href.epa)

#alternative plot for first cat
plot(kernel.href.bivar[[1]])
plot(kernel.href.epa[[1]])

#UD data
kernel.href.bivar[[1]]@data

#h value for bandwidth
kernel.href.bivar[[2]]@h
kernel.href.bivar[[2]]@h$h

#least-squares cross validation for h, this will yield a warning of convergence, can't converge on solution so can't use 
kernel.lscv.bivar <- kernelUD(panthers[,"CatID"], h="LSCV", kern="bivnorm")

#manually adjust h, the bandwith parameter, "hard coding"
kernel.bivar.h1000 <- kernelUD(panthers[,"CatID"], h=1000, kern="bivnorm")

```

```{r, include=FALSE, message=F, warning=F, echo=FALSE}
## C. Local convex hull

## 14 points seems to be a good choice (rough asymptote for all animals)
## the k-LoCoH method:
nn <- LoCoH.k(panthers[,1], k=14)

## Graphical display of the results
plot(nn, border=NA)

length(nn)
names(nn)

lch_panther_all=data.frame()
for(i in 1:length(nn)){
  iso=as.data.frame(nn[[i]])
  iso=subset(iso, percent<100)
  iso=which.min(abs(iso$percent-95))
  
  lch=(nn[[i]][iso,])
  lch=fortify(lch)
  lch$id=names(nn)[i]
  lch_panther_all=rbind(lch, lch_panther_all)
}

lch_panther_all_holes=subset(lch_panther_all, hole==T)

lch_panther_all=subset(lch_panther_all, hole==F)

(lch.plot <- ggplot()+
geom_polygon(data=lch_panther_all, aes(x=long, y=lat, group=group, fill=id), alpha=.8)+
geom_polygon(data=lch_panther_all_holes,aes(x=long, y=lat, group=group), fill="white", alpha=1)+
    labs(fill = "Panther ID",
         x = "Easting (m)",
         y = "Northing (m)",
         title = "Local Convex Hull")+
    ggthemes::theme_few()+
  scale_x_continuous(label=scientific,
                     breaks= c(600000, 725000),
                     limits = c(600000, 725000) )+
scale_y_continuous(label=scientific, 
                   breaks=c(200000,300000,400000),
                   limits = c(200000,400000)))

```

```{r, include=FALSE, message=F, warning=F, echo=FALSE}
## D. Brownian Bridge

# need a time series to make this work bc random walk through time
#Re-format Juldate information:
#function for taking characters of a string from rightmost value
substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

#re-format
panthers$Juldate <- as.character(panthers$Juldate)
panther.date <- as.numeric(substrRight(panthers$Juldate, 3))
panthers$Date <-as.Date(panther.date, origin=as.Date("2006-01-01"))

#convert to POSIXct object
panthers$Date <- as.POSIXct(panthers$Date,"%Y-%m-%d", tz = "EST")

#convert to data frame
panther.df <- as.data.frame(panthers)

#make trajectory object
panther.ltraj <- as.ltraj(xy=coordinates(panthers), date=panthers$Date, id=panthers$CatID, typeII=T)

#inspect
head(panther.ltraj)
head(panther.ltraj[[1]], 2)
summary(panther.ltraj)

#plot, how much room is there to move across landscape 
plot(panther.ltraj)

#telemetry error (average)
sigma2 <- 450

#estimate sigma1, individual- specific metric of pace of movement, will vary
sigma1 <- liker(panther.ltraj, sig2 = sigma2, rangesig1 = c(2, 100))

#inspect
sigma1

#brownian bridge for Cat 147
 bb.147 <- kernelbb(panther.ltraj[6], sig1 = 7.2, sig2 = sigma2, grid = 200)

#all panthers
sig1 <- c(sigma1[[1]]$sig1, sigma1[[2]]$sig1, sigma1[[3]]$sig1, sigma1[[4]]$sig1, sigma1[[5]]$sig1, sigma1[[6]]$sig1)

bb.panther <- kernelbb(panther.ltraj, sig1 = sig1, sig2 = sigma2, grid = 200)

#plot
plot(panther.ltraj[6])
plot(bb.147)
```

 
```{r, include=FALSE, message=F, warning=F, echo=FALSE}
## Contrasting and making the 95%

#home range area estimates
kernel.95 <- getverticeshr(kernel.href.bivar, percent=95)

(bivar.k.plot <- ggplot()+
  geom_polygon(kernel.95, mapping =aes(x=long, y=lat, group=group, fill=id), alpha = 0.8)+
  labs(x = "Easting (m)",
       y = "Northing (m)",
       fill = "Panther ID",
       title = "Bivariate Kernel")+
  scale_x_continuous(label=scientific,
                     breaks= c(600000, 725000),
                     limits = c(600000, 725000) )+
scale_y_continuous(label=scientific, 
                   breaks=c(200000,300000,400000),
                   limits = c(200000,400000))+
    ggthemes::theme_few())


bb.95 <- getverticeshr(bb.panther, percent=95)

(bb.95.plot <- ggplot()+
  geom_polygon(bb.95, mapping =aes(x=long, y=lat, group=group, fill=id), alpha = 0.8)+
  labs(x = "Easting (m)",
       y = "Northing (m)",
       fill = "Panther ID",
       title = "Brownian Bridge")+
  scale_x_continuous(label=scientific,
                     breaks= c(600000, 725000),
                     limits = c(600000, 725000) )+
scale_y_continuous(label=scientific, 
                   breaks=c(200000,300000,400000),
                   limits = c(200000,400000))+
    ggthemes::theme_few())


#contrast area
mcp95$area
kernel.95$area
bb.95$area

#plot
par(mfrow=c(1,2))  #sets up the graph window to store two graphs
plot(land_sub)
plot(kernel.95, col=kernel.95$id)

plot(land_sub)
plot(mcp95, add=TRUE, col=kernel.95$id)
```

```{r, message=F, warning=F, echo=FALSE, fig.dim=c(12, 12)}
cowplot::plot_grid(mch.plot, bivar.k.plot, lch.plot, bb.95.plot, nrow = 2, ncol = 2, labels = "AUTO")
```

# 6. 
If I had to present home range of the panthers, I would use the results from the Brownian Bridge model because it uses both space and time in addition to movement trajectories.

# 7. 
The difference between Design II and Design III is that in Design II, population - level data is incorporated in that use by individuals is quantified and compared to availability data taken at the population level and so is not specific to each individual. But, in Design III, both use and availability data are quantified for each individual. 

# 8.

Line 293 is using dcast from {reshape} to select only the cat ID and land cover, and then counting the occurrences of each land cover type, per cat (AKA the "length" argument) with the columns from 2:n being the different land cover classes. 
```{r, include=FALSE, message=F, warning=F, echo=FALSE}

#8.3.4 Resource Selection
########################################################

###########################################
#8.3.4.1 Point selection functions
###########################################

#use data
use <- raster::extract(layers, panthers)
use <- data.frame(use)

#inspect
head(use)
str(use)

#add CatID
use$CatID <- as.factor(panthers$CatID)

#reformat: this is line 293
useCatID <- dcast(use, CatID~landcover, length, value.var="CatID")

# try with dplyr grouping to see if I get the same results
# use_counts <- use %>%
#                 select(CatID, landcover) %>%
#                 group_by(CatID, landcover) %>%                 summarise(n()) %>%
#                 ungroup() %>%
#                 pivot_wider(names_from = "landcover", values_from = "n()", names_sort = T)

#inspect
useCatID

#add land-cover names
newclass.names <- unique(classification[,3:4])
names(useCatID) <- c("CatID", as.character(newclass.names[1:13,2]))

#inspect
useCatID
```

# 9. Design resource selection models 

```{r, include=FALSE, message=F, warning=F, echo=FALSE}

## Design II availability: population availability
#get availability points
set.seed(8)
rand.II <- sampleRandom(layers, size=1000)
rand.II <- data.frame(rand.II)

#inspect
head(rand.II)
str(rand.II)

rand.II.land <- as.factor(rand.II$landcover)

#get counts of each landcover type
avail.II <- tapply(rand.II.land, rand.II.land, length)

#inspect
avail.II

#add land-cover names
names(avail.II) <- as.character(newclass.names[1:14,2])

#inspect
avail.II

#remove exotics, which were not observed in use sample
avail.II <- avail.II[c(-14)]
```


```{r, include=FALSE, message=F, warning=F, echo=FALSE}
## Design III: within home-range availability for each individual

cat.unique <- unique(panthers$CatID)
samples <- 200
rand.III <- matrix(nrow=0, ncol=4)

#loop for all individuals
for(i in 1:length(cat.unique)){

  id.i <- cat.unique[i]
  cat.i <- panthers[panthers$CatID==id.i,]
  mcp.i <- mcp(SpatialPoints(coordinates(cat.i)), percent = 99)
  rand.i <- spsample(mcp.i, type="random", n=samples)
  rand.i.sample <- raster::extract(layers, rand.i)

  #make a matrix of CatID and rand samples
  cat.i <- as.numeric(rep(cat.unique[i], length(rand.i)))
  rand.cat.i <- cbind(cat.i, rand.i.sample)
  rand.III <- rbind(rand.III, rand.cat.i)
}

#inspect
head(rand.III)
class(rand.III)
str(rand.III)

#reshape data
rand.III <- data.frame(rand.III)
rand.III$cat.i <- as.factor(rand.III$cat.i)
avail.III <- dcast(rand.III, cat.i~landcover, length, value.var="cat.i")

names(avail.III)[2:14] <- as.character(newclass.names[1:13,2])
#inspect
avail.III

```


```{r, include=FALSE, message=F, warning=F, echo=FALSE}
## Selection ratios

#Design II:
sel.ratioII <- widesII(u=useCatID[,c(2:ncol(useCatID))], a=as.vector(avail.II), avknown=FALSE, alpha = 0.05)

sel2use <- sel.ratioII[["used.prop"]]

sel2use.df <- as.data.frame(sel2use)

#inspect
sel.ratioII
sel.ratioII$wi
sel.ratioII$se.wi

#plot
#plot(sel.ratioII, errbar = c("CI"))

# selection comaparisons
selections <- as.data.frame(sel.ratioII[["comparisons"]])
selections.sig <- 
  

#Design III:
sel.ratioIII <- widesIII(u=useCatID[,c(2:ncol(useCatID))], a=avail.III[,2:14], avknown=FALSE, alpha = 0.05)

#inspect
sel.ratioIII
sel.ratioIII$wi
sel.ratioIII$se.wi
sel.ratioIII$ICwiupper
sel.ratioIII$ICwilower

#plot
#plot(sel.ratioIII, errbar = c("CI"))
```

## a. 
Based on the output of `widesII`, specifically the sel2use plot, all panthers are not using all habitats identically. There are some habitats that show use by multiple cats (for example, cypress swamp has 2 cats using similarly) but not one that has all 6 using the same. 

## b. 
Based on the outputs of `sel.ratioII$wi`, habitats that show a positive selection are barrenland, dry prarie, cypress swamp, pinelands, hardwood swamp, and upland forest.

The classes that show a significant positive association (based on upper confidence intervals above 1) are coastal wetland, pasture/grassland, barren land, dry prairie, freshwater marsh, cypress swamp, pine lands, hardwood swamp, and upland forest. 

```{r, include=FALSE, message=F, warning=F, echo=FALSE}
sel.ratioII$wi

sel.ratioII$ICwiupper
```

## c. 
Based on the output of `sel.ratioIII$wi`, habitats that show a positive selection are all except cropland and pasture/grassland, which would be the negative associations.
```{r, include=FALSE, message=F, warning=F, echo=FALSE}
sel.ratioIII$wi

sel.ratioIII$ICwilower

```

## d. 
The differences between the Design II and III global results are that the design III says that many more habitats were positively selected, even habitats that you would not expect panthers to use. This is probably because Design III is on the individual level and so if individuals are using habitats that the rest are not, it would still be counted. 

