---
title: "ECOL 620 Lab 4"
author: "Natalie Schmer"
date: "2/11/2021"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r, message = F}
#load packages
{
  library(raster)            
  library(rasterVis)         
  library(landscapemetrics)  
  library(rgdal)
  library(tidyverse)
  library(here)
}

# pull up all the possible metrics 
possible_mets <- list_lsm()
```

# Fort Collins Raster
## 1. Bar plot of land cover types 
```{r, warning=F, message=FALSE, fig.dim=c(12,8)}
nlcd <- raster(here("data/data_for_lab4/fort_collins.tif"))
nlcd <- as.factor(nlcd)

#add names of categories to raster layer
land_cover <- levels(nlcd)[[1]]

land_cover[,"landcover"] <- c("Open Water", "Developed, Open Space","Developed, Low Intensity",
                              "Developed, Medium Intensity","Developed, High Intensity",
                              "Barren Land","Deciduous Forest", "Evergreen Forest","Mixed Forest",
                              "Shrub/Scrub","Grassland/Herbaceous","Pasture/Hay","Cultivated Crops",
                              "Woody Wetlands","Emergent Herbaceous Wetlands")

levels(nlcd) <- land_cover

#plot
# this is the color scheme, in order  
land_col <- c("#4f6d9f", "#decece", "#d29b85", "#de3021", "#9d1f15",
              "#b2afa5", "#7aa76d", "#336338", "#c0cb99","#cebb89", "#edecd0",
              "#ddd75c", "#a67538", "#bfd7eb", "#7ba3be")

#plot with rasterVis
#levelplot(nlcd, col.regions=land_col, xlab="", ylab="")

# Bar plot 
# quantify % land cover of each type and join with IDs to have the cover types
data_for_percent_land_FC <- lsm_c_pland(nlcd)

colnames(data_for_percent_land_FC)[3] <- "ID"

data_for_percent_land_FC <- merge(land_cover, data_for_percent_land_FC, by = "ID")

data_for_percent_land_FC$color_codes <- land_col

# Reorder according to order of nlcd and plot
 data_for_percent_land_FC %>% 
        mutate(landcover = fct_relevel(landcover, "Open Water", "Developed, Open Space","Developed, Low Intensity",
                              "Developed, Medium Intensity","Developed, High Intensity",
                              "Barren Land","Deciduous Forest", "Evergreen Forest","Mixed Forest",
                              "Shrub/Scrub","Grassland/Herbaceous","Pasture/Hay","Cultivated Crops",
                              "Woody Wetlands","Emergent Herbaceous Wetlands")) %>% ggplot()+
  geom_bar(mapping = aes(x = landcover, y = value), fill = land_col, stat = "identity") +
  theme_classic()+
  labs(x= NULL,
       y = "PercentLand Cover Type (%)")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 12))
```

# Calculating metrics for open water 
## 2. 

```{r, warning=F, message=FALSE}
#4-neighbor rule
four_patch_rule <- lsm_c_np(nlcd, directions = 4)  

# Pull out specific value for class 11 aka open water
openwp4 <- four_patch_rule[four_patch_rule$class==11,]$value  

#8-neighbor rule
eight_patch_rule <- lsm_c_np(nlcd, directions = 8) 

openwp8 <- eight_patch_rule[eight_patch_rule$class==11,]$value  
```
The number of open water patches in the greater Fort Collins region using the 4-neighborhood rule: `r toString(openwp4)` and the 8-neighborhood rule: `r toString(openwp8)`

## 3.  

```{r, warning=F, message=FALSE}

patch_metrics_3 <- calculate_lsm(nlcd, 
                                level = "patch", 
                                metric = "area", 
                                directions = 4, 
                                full_name = T) 

lg_owp_3 <- patch_metrics_3 %>% 
              dplyr::filter(class == "11") 

lg_owp_v <- max(lg_owp_3$value)

```
The largest open water patch size in hectare using the 4-neighborhood rule is `r toString(lg_owp_v)`.

## 4.  
```{r, warning=F, message=FALSE}
# hectacres check 
mean_owp_v <- mean(lg_owp_3$value)

# in km^2
mean_owp_v_km <- mean_owp_v*0.01

```
Mean open water patch size  using the 4-neighborhood rule is `r toString(mean_owp_v)` hectares or `r toString(mean_owp_v_km)` km^2.

## 5.   
```{r, warning=F, message=FALSE}

patch_metrics_8 <- calculate_lsm(nlcd,
                                 level = "patch", 
                                 what = "lsm_p_core",
                                 directions = 8, 
                                 edge_depth = 3,
                                 full_name=T)

corea_owp_8 <- patch_metrics_8 %>% 
              dplyr::filter(class == "11") 

q5_range <- range(corea_owp_8$value)
```
The range of core area open water patches in hectares using the 8-neighborhood rule and
edge depth of 3 is `r toString(q5_range)`.


## 6. Correlations 

Of the patch-based metrics (using the 4-neighborhood rule), which has the strongest
negative correlation? Include the top-three pairings.

I only included correlations that had the absolute value of the coefficient of < 90. The strongest negative correlation is between para (perimeter-area ratio) and circle (related circumscribing circle) (-0.825). From reading the package documentation, this seems to mean that  the less circular a shape is (low perimeter area ratio), the larger the circle is, meaning the patch is not very compact. 

The top three pairings (using absolute values) were: 
1. Frac (fractal dimension index, describes patch complexity) and circle (0.865) 
2. Shape (shape index, ratio of actual vs hypothetical minimum patch perimeter) and gyrate (radius of gyration, the distance from each cell to the patch centroid) (0.846). 
3. Shape and frac (0.837)
```{r, warning=F, message=FALSE}
# only taking correlations with absolute value of less than 90
q6 <- calculate_lsm(nlcd, 
                    level = "patch", 
                    directions = 4,
                    full_name=T)

corr <- show_correlation(data = q6[,1:6], method = "pearson", label=T) 

corr_data <- as.data.frame(corr$data)

# remove the 1 becuase those don't count 
corr_data_1 <- corr_data %>% 
                mutate(abs_value = abs(value)) %>% 
                filter(abs_value < 1)
                
```

## 7. Plot showing the relationship between the number of open water patches using the 4-neighborhood rule and grain size of 30m, 60m, 120m, 240, and 600m.

As the resolution decreases due to increased aggregation factors, there is less detail in the resulting raster (other than i = 1, or resolution staying 30) and the number of patches decreases. 

```{r, warning=F, message=FALSE}
#reload nlcd for 7 
q7_nlcd <- raster(here("data/data_for_lab4/fort_collins.tif"))

#Make dataframe to store loop components
q7_data <- data.frame() %>% 
            mutate(aggregate_factor = NA,
                   patch_number = NA)
z <- data.frame(NA)

#Create the sequence, define steps and where to put data 

for (i in as.numeric(c((30/30), (60/30), (120/30), (240/30), (600/30)))) {
  
  print(i)
  
  new_raster <- raster::aggregate(q7_nlcd, fact = i, fun = "modal")
  
  patch_metrics <- calculate_lsm(new_raster,
                                 level = "class", 
                                 what = "lsm_c_np",
                                 directions = 4,  
                                 full_name = T) 
  
  z$patch_number <- patch_metrics %>% 
    dplyr::filter(class == "11") %>% 
    select(value) %>% 
    as.numeric()
 
  z$aggregate_factor <- i
  z$grain_size <- z$aggregate_factor*30
  
  q7_data <- rbind(z, q7_data)
}

q7_data %>% 
              select(-NA.) %>% 
  ggplot(., aes(grain_size, patch_number))+
  geom_point()+
  geom_line()+
  ggthemes::theme_clean()+
  labs(x = "Grain Size",
       y = "Number of Open Water Patches")

```

# Reclassification 
```{r, warning=F, message=FALSE}
#land-cover type (nlcd original categories)
#1 = forest:41-43
#2 = developed:21-24
#3 = agriculture:81,82
#4 = herbaceous:71-74
#5 = open:31, 51-52
#6 = wetland:90,95
#7 = water:11-12

# set up reclassifications 
nlcd <- raster(here("data/data_for_lab4/fort_collins.tif"))

nlcd <- as.factor(nlcd)

reclass_raster <- nlcd
reclass_ids <- as.data.frame(reclass_raster@data@attributes)

values(reclass_raster) <- 0 

# change the covers 
reclass_ids <- reclass_ids %>% 
                  mutate(reclass_land = case_when(ID %in% c(41:43) ~ "forest",
                                                  ID %in% c(21:24) ~ "developed",
                                                  ID %in% c(81:82) ~ "agriculture",
                                                  ID %in% c(71:74) ~ "herbaceous",
                                                  ID %in% c(31, 51:52) ~ "open",
                                                  ID %in% c(90, 95) ~ "wetland",
                                                  ID %in% c(11:12) ~ "water")) %>% 
                  mutate(reclass_id = case_when(reclass_land == "forest" ~ 1,
          reclass_land == "developed" ~ 2,
          reclass_land == "agriculture" ~ 3,
          reclass_land == "herbaceous" ~ 4,
          reclass_land == "open" ~ 5,
          reclass_land == "wetland" ~ 6,
          reclass_land == "water" ~ 7)) %>% 
  mutate(reclass_id = as.numeric(reclass_id)) %>% 
                  dplyr::select(ID, reclass = reclass_id)

# make a matrix 
reclass_matrix <- as.matrix(reclass_ids)

# new levels df
nlcd.levels <- levels(nlcd)[[1]]

# join IDs and new IDs
reclass.mat <- cbind(levels(nlcd)[[1]], reclass_matrix)

# Check 
reclass.mat

# Make it into a matrix 
reclass.mat <- as.matrix(reclass.mat)

#only want the original ID and the new ID
reclass.mat <- reclass.mat[,c(1, 3)]
reclass.mat

#check class
class(reclass.mat[,2])

#reclass.mat[,3] <- as.integer(reclass.mat[,3])
#reclass.mat[,1] <- as.integer(reclass.mat[,1])

# Reclassify
reclass_raster <- reclassify(nlcd, reclass.mat)

# Plot to check 
plot(reclass_raster)

# Looks ok, open water makes sense and has a large developed patch that seems to match the city 
```

## 8. 

```{r, warning=F, message=FALSE}
q8_4 <- lsm_c_np(reclass_raster, directions = 4) 

value_4 <- q8_4 %>% 
                filter(class == 3) %>% 
                select(value)

q8_8 <- lsm_c_np(reclass_raster, directions = 8)

value_8 <- q8_8 %>% 
                filter(class == 3) %>% 
                select(value)
```
The 4-neighborhood rule gives `r toString(value_4)` agriculture patches and the 8-neighborhood rule gives `r toString(value_8)` agriculture patches. 

## 9.  

```{r, warning=F, message=FALSE}
q9_4 <- calculate_lsm(reclass_raster, 
                      level = "patch",
                      metric = "enn", 
                      directions = 4, 
                      full_name=T) 

value_4_9 <- q9_4 %>% 
                dplyr::filter(class == 3) 
                
mean_q9 <- mean(value_4_9$value, na.rm = T)
```
The mean euclidean nearest neighbor distance using the 4 neighborhood rule for agriculture patches is `r toString(mean_q9)` m.

## 10. Plot of developed and open water landcover with number of core areas. 

```{r, warning=F, message=FALSE}
# p = patches, ncore = number of cores
show_lsm(reclass_raster, 
         what ="lsm_p_ncore",
         class = c(2, 7),
         directions = 4,
         edge_depth = 3)

```

# Calculating other metrics for reclassified fort Collins

## 11. Land-cover richness describes the number of different types of land cover. In the case of the reclassified rather, there are 7 different types.

```{r}
richness <- length(unique(values(reclass_raster)))
richness
```

## 12. Shannon diversity is a metric that takes into account number/richness and evennes of land cover types. A higher diverity index indicates less fragmentation or fewer edges and patches, but is also better interpreted when comapring to values for other landscapes. 
```{r}
table(values(reclass_raster))

C <- table(values(reclass_raster))
P <- C / sum(C)
D <- -sum(P * log(P))
```
Shannon diversity index here is `r toString(D)`.

## 13. Shannon evennessdescribes distribution of relative areas among land- cover types. A higher eveness value means greater landscape diversity.  
```{r}
E <- D/log(length(C))
```
Shannon evenness here is `r toString(E)`

## 14. Contagion is one of the aggregation metrics and is based on cell adjacencies and describes the probability of 2 random cells belonging to the same class. The contagion for the reclassified raster is 45.3, with higher values up to 100 meaning all cells are equally adjacent to all the other classes and lower values down to 0 meaning all cells are unevenly distributed.  
```{r}
lsm_l_contag(reclass_raster)
```

