---
title: "ECOL 620 Lab 2"
author: "Natalie Schmer"
date: "1/28/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
save.image(file = "/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab2.RData")

load("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab2.RData")
```

# Load packages 
```{r}
{
  library(raster)      #for raster data; version 2.6-7 used
  library(rgdal)       #for raster data, projections; version 1.3-4 used
  library(rgeos)       #for buffer analysis; version 0.3-28 used
  library(here)
  library(tidyverse)
  }
```

# 1. Describing functions
- `crop()` from {raster} uses an extent of one object to geographically subset another object.
- `extent()` from {raster} gives the extent of a spatial object with xmin/ xmax and ymin/ ymax.
- `disaggregate()` from {raster} creates a raster with a higher resolution due to smaller cells, retains the values as the original raster unless using the bilinear method. 
- `aggregate()` from {raster} creates a raster with a lower resolution due to larger cells.  
- `cellStats()`
- `mask()`
- `res()`
- `gBuffer()`
- `rpois()`

# 2. Creating a 20 x 20 raster with poisson distribution, lambda = 5. 
```{r}
# Create raster
q2rast <- raster(ncol = 20, nrow = 20, xmn = 0, xmx = 20, ymn = 0, ymx = 20)

# Assign poisson dist to fill the raster cells
q2rast[] <- rpois(ncell(q2rast), lambda=5)

# plot
plot(q2rast, axes = F, box = F)
text(q2rast, digits = 2)
res(q2rast) 

# Stats 
(mean <- cellStats(q2rast, mean))  
(variance <- cellStats(q2rast, var))
```

# 3. Aggregate the CO elevation 
```{r}
# Load elevation raster 
co_elev <- raster(here("data/data_for_lab2/elevation/srtm_15_05.tif"))

# For loop to aggregate raster, calc variance, get resolution, and make into a df to plot from 
new_data <- data.frame()
for (i in as.numeric(seq(5, 255, by = 25))) {
  print(i)
  new_raster <- raster::aggregate(co_elev, fact = i, fun = mean)
  variance <- cellStats(new_raster, var)
  res_new_raster <- res(new_raster)[1]
  z <- c(aggregate_factor = i, var = variance, res =  res_new_raster)
  new_data <- rbind(z, new_data)
}

# For some reason the column names did not carry over to the new data frame 
# Look at the components we assigned in the loop 
z
new_data

# The cols and values in z match the 255 factor row of new_data so just rename 
names(new_data)
new_data <- new_data %>% 
                      rename(aggregate_factor = "X5", 
                             var = "X303795.930365646", 
                             res = "X0.00416666666666667")

```





