---
title: "ECOL 620 Lab 1"
author: "Natalie Schmer"
date: "1/21/2021"
output: word_document
editor_options: 
  chunk_output_type: console
---
# Load in packages 
```{r, message=F, warning=F}
{
  library(tidyverse)
  library(rgdal)
  library(viridis)
  library(tidylog)
  library(ggthemes)
  library(ggrepel)
  library(here)
  library(sp)
  library(cowplot)
}

#need to download old usmap 
install.packages("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/usmap_0.5.1.tar", repos = NULL, type="source")
```

# 1. Briefly describe (1-2 sentences) what each of the following functions achieve.

- `c()` is for concatenate, and combines the arguments into a list or vector.
- `geom_smooth()` is in {ggplot2} and draws a line through data to help show the pattern of the data, helps a lot when there is a lot on the plot. Need to specify the formula and fit arguments.
- `rep()` replicates values as specified in the arguments.
- `filter()` in {dplyr} subsets dataframe rows using values in columns that satisfy the given arguments.
- `%>%()` is a pipe! From {magrittr}, moves whatever object is on the left to the operation on the right.
- `png()` writes an object (like a plot) to a .png file. Need to specify file path as an argument and can also set height, width, other aesthetics.
- `coord_map()` is for map projection in {ggplot2}.
- `plot_grid()` from {cowplot} allows for arranging multiple plots in a grid. Can specify number of rows, columns, adjust alignments, labels with a variety of arguments.

# US Maps
## 2. Ploting the contiguous U.S. states colored by area 
```{r, message = F, warning = F}
# Load the data we used earlier, join states to df with areas
states <- as.data.frame(state.x77)
states$region <- tolower(rownames(states))
states_map <- map_data("state")
fact_join <- left_join(states_map, states, by = "region")

(state_areas <- ggplot(fact_join, aes(long, lat, group = group))+
  geom_polygon(aes(fill = Area), colour = "white")+
  scale_fill_viridis_c(option = "D")+
  ggthemes::theme_clean() +
  coord_map("bonne", lat0 = 40)+
  labs(y = "Latitude", 
       x = "Longitude", 
       fill = "Area (sq. miles)",
       title = "Area of US States")+
  theme(legend.position = "bottom",
        legend.text = element_text(angle = 90, vjust = 0.5))
)
```

## 3 Three additional plots of the same data, each with different color scale and geographic projection, put together with `plot_grid` function to make a three-panel horizontal plot.
```{r, message = F, echo = T, fig.show="hide"}

#1 
(state_areas_1 <- ggplot(fact_join, aes(long, lat, group = group))+
  geom_polygon(aes(fill = Area), colour = "white")+
  scale_fill_viridis_c(option = "A")+
  ggthemes::theme_clean() +
  coord_map("mercator")+
  labs(y = "Latitude", 
       x = "Longitude", 
       fill = "Area (sq. miles)",
       title = "Area of US States, \nMercator Projection")+
  theme(legend.position = "bottom",
        legend.text = element_text(angle = 90, vjust = 0.5))
)

#2
(state_areas_2 <- ggplot(fact_join, aes(long, lat, group = group))+
  geom_polygon(aes(fill = Area), colour = "white")+
  scale_fill_viridis_c(option = "E")+
  ggthemes::theme_clean() +
  coord_map("stereographic")+
  labs(y = "Latitude", 
       x = "Longitude", 
       fill = "Area (sq. miles)",
       title = "Area of US States, \nStereographic Projection")+
  theme(legend.position = "bottom",
        legend.text = element_text(angle = 90, vjust = 0.5))
)

#3
(state_areas_3 <- ggplot(fact_join, aes(long, lat, group = group))+
  geom_polygon(aes(fill = Area), colour = "white")+
  scale_fill_viridis_c(option = "C")+
  ggthemes::theme_clean() +
  coord_map("cylequalarea", lat0 = 40)+
  labs(y = "Latitude", 
       x = "Longitude", 
       fill = "Area (sq. miles)",
       title = "Area of US States,\nCylequalarea Projection")+
  theme(legend.position = "bottom",
        legend.text = element_text(angle = 90, vjust = 0.5))
)
```

### 3-panel of the different color scales and projections
```{r, fig.dim = c(12,6), message = F}
plot_grid(state_areas_1, state_areas_2, state_areas_3, nrow = 1)
```

# US University Data
## 4. Plot Colorado colleges and universities within the top 96th quantile of total student enrollment, add in `geom_text_repel(data=NAME_OF_YOUR_FILTERED_DATA, aes(label=NAME, x=LON, y = LAT), force=20, size=2)` 
```{r, message = F, warning = F}
# Read in the csv of university data  
colo_unis <- read.csv(here::here("data/data_for_lab1/Colleges_and_Universities/colorado_universities.csv")) 

# Filter to only CO 
colo_unis <- colo_unis %>% 
                filter(LSTATE == "CO")
# check 
unique(colo_unis$LSTATE)

# Calculate and filter the 96th quantile universities
(quan_96 <- quantile(colo_unis$TOT_ENROLL, .96, na.rm = T))

co_96 <- colo_unis %>% 
                filter(TOT_ENROLL >= quan_96)

# Load in the CO shapefile 
co_counties <- readOGR(here::here("data/data_for_lab1/counties/Colorado_County_Boundaries.shp"))

# Set the min and max enroll
min_enroll <- min(colo_unis$TOT_ENROLL)
max_enroll <- max(colo_unis$TOT_ENROLL)

# Plot 
ggplot() +
  geom_polygon(data = co_counties, 
               aes(x = long, 
                   y = lat, 
                   group = group), 
               fill = NA, 
               color ="black", 
               lwd =.1) +
  geom_point(data = co_96, 
             aes(x = LON, 
                 y = LAT, 
                 size = TOT_ENROLL, 
                 colour = TOT_ENROLL), 
             alpha =.9) +
  coord_map("bonne", lat0 = 40)+
  theme_bw()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())+
  theme(panel.border=element_blank())+
  scale_color_viridis(limits = c(min_enroll, max_enroll),
                      breaks = seq(5000, 30000, by=5000), 
                      name = "Total\nenrollment")+
  guides(color= guide_legend(), size=guide_legend())+
  scale_size_continuous(limits = c(min_enroll,max_enroll), 
                        breaks = seq(5000, 30000, by=5000),
                        name = "Total\nenrollment") +
    geom_text_repel(data = co_96, 
                    aes(label = NAME, 
                        x = LON, 
                        y = LAT), 
                    force = 20, size = 2)+
  theme_clean()+
  labs(y = "Latitude", 
       x = "Longitude",
       title = "CO Universities within the top\n96th quantile of student enrollment")
```

## 5. Other universities: On, Wisconsin!
Wisconsin county boundary shapefile downloaded from https://data-wi-dnr.opendata.arcgis.com/datasets/county-boundaries-24k 
```{r, warning=FALSE, message=F}
# load in shapefiles for WI counties and universities
wi_counties <- readOGR(here::here("data/County_Boundaries_24K-shp/County_Boundaries_24K.shp"))

us_uni <- readOGR(here::here("data/data_for_lab1/Colleges_and_Universities/CollegesUniversities.shp"))

# Filter out WI schools
wi_uni <- subset(us_uni, LSTATE=="WI")

# check crs 
proj4string(wi_counties)
proj4string(wi_uni)

# reproject counties with the uni projection 
wi_counties_reprj <- spTransform(wi_counties, "+proj=longlat +datum=WGS84 +no_defs")

# make wi unis a df 
wi_unis_df <- as.data.frame(wi_uni)

#remove ins_size =<0 as those are not reported or not applicable
wi_unis_df <- wi_unis_df %>% 
              dplyr::filter(INST_SIZE > 0) 
```

### WI Plot
```{r, warning=FALSE, message=F}
ggplot() +
  geom_polygon(data = wi_counties_reprj, 
               aes(x = long, 
                   y = lat, 
                   group = group),
               fill = NA, color ="black", lwd=.1) +
  geom_point(data = wi_unis_df %>% filter(NAME != "University of Wisconsin-Madison"), 
               aes(x = LON, 
                   y = LAT, 
                   size=INST_SIZE, 
                   colour=INST_SIZE), 
              alpha=.7) + 
    geom_point(data = wi_unis_df %>% filter(NAME == "University of Wisconsin-Madison"), 
               aes(x = LON, 
                   y = LAT, 
                   size=INST_SIZE), 
               alpha=.6, color = "blue", show.legend = F) +
  coord_map("bonne", lat0 = 40)+
  theme_clean()+
  theme(panel.grid.minor=element_blank(),
        panel.grid.major=element_blank())+
  theme(panel.border=element_blank())+
  scale_color_viridis(limits = c(1, 5),
                      labels = c("Under 1000",
                                 "1000 - 4999",
                                 "5000 - 9999",
                                 "10000 - 19999",
                                 "20000 and above"),
                      name = "Institution\nsize") +
  guides(color= guide_legend(), size=guide_legend()) +
  scale_size_continuous(limits=c(1, 5),
                        breaks=seq(1, 5, by= 1),
                        labels = c("Under 1000",
                                 "1000 - 4999",
                                 "5000 - 9999",
                                 "10000 - 19999",
                                 "20000 and above"),
                        name = "Institution\nsize")+
  labs(y = "Latitude", 
       x = "Longitude",
       title = "University of Wisconsin-Madison")

```

