---
title: "ECOL 620 Lab 6"
author: "Natalie Schmer"
date: "2/25/2021"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r, message = F}
#load packages
library(raster)           
library(rgdal)            
library(rgeos)            
library(gdistance)        
library(igraph)           
library(fitdistrplus)     
library(fdrtool)          
library(tidyverse)
library(viridis)
library(landscapemetrics) 

# rds of distances from the modeling
all.dist <- readRDS("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab6/all.dist.rds")
```

# Formatting data
```{r, message = F, fig.show="hide", include=F}
land <- raster("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab6/panther_landcover")

#inspect
projection(land)
res(land)

#label projection for later use
crs.land <- projection(land)
plot(land)

#public areas in need of connections
public <- readOGR("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab6/panther_publicland.shp")
projection(public)
projection(public) <- crs.land
names(public@data)                        #attributes table
head(public@data)                         #attributes table

#get the centroids of plots
public_centroids <- gCentroid(public, byid=T)
public_centroids@coords
public_centroids@coords[1,]  

#import reclassification table
classification <- read.table("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab6/resistance reclass.txt", header=T)

#inspect
head(classification,3)

#reclass
class <- as.matrix(classification[,c(1,3)])
land_cost <- reclassify(land,rcl=class)


#create a conductance transition layer: inverse of resistance data
land_cond <- transition(1/land_cost, transitionFunction=mean, 8)

#make correction; type=c for lcps; type = r for circuit (identical results for this example, so just use c)
land_cond <- geoCorrection(land_cond, type="c", multpl=F)
```


# 1. Methods
Euclidean distance is the most direct distance between points and does not take into account the resistance of the landscape. Least-cost distance is also a measure of the shorter possible distances between points but starts to take into account landscape resistances and calculates the cumulative cost of different landscape characteristics to determine a path. Circuit theory looks at the redundancy in paths to determine the most optimal commuting path based on random walks, which account for alternate paths. Randomized shortest path is similar to circuit theory and least cost distance while assuming that animals don't know the optimal path and so might be wandering to find the best path. Rsp also uses the "theta" parameter to tune the model probabilities for the least-cost path. 

I think the most robust estimation would come from the randomized shortest path since there is the additional option to control theta values in addition to taking into consideration landscape resistance and the least cost path. 

# 2. 
## a. 
```{r, message = F, message=F, warning=F}
length(unique(values(land)))
```
There are 42 unique land cover classes in the original panther land cover raster. 

## b. 
```{r, message = F, message=F, warning=F}
#import reclassification table
classification <- read.table("/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab6/resistance reclass.txt", header=T)

#inspect
head(classification,3)

#reclass
class <- as.matrix(classification[,c(1,3)])
land_cost <- reclassify(land,rcl=class)

(length(unique(land_cost)))
```
After reclassification, there are 10 classes. 

## c. From the supplied rankings, class 11 or coastal wetland would provide the most resistance and class 1 or hardwood swamp would provide the least resistance. 


## d. Resistance measures are determined by looking at data that gives insight into things like species distribution, habitat use, and movement, and a numerical value is assigned to raster classes based on how difficult it would be for an animal to pass through a class. Can also be calculated with the inverse of permeability if permeability is provided. 


# 3. Correlations 
```{r, message = F, message=F, warning=F}
(round(cor(all.dist),3))
```
The most correlated are rSP and lcd (0.996) and the least correlated are euclidean and circuit (0.854). 

# 4.
In plot 9.7 of the Fletcher text, a single point represents the scaled effective distances (mean of 0 and variance of 1) between the start and end points for a given pair of distance metrics (Euclidean, least cost, circuit, rSP). There are 10 points because there are 10 distance measures that come out of whatever distance function you apply (`pointDistance`, `costDistance`, `commuteDistance`, `rSPDistance`)

# 5. 
```{r,message = F, message=F, warning=F, fig.show="hide"}
#public land web
projection(public)
projection(public) <- crs.land
names(public@data)                        #attributes table
head(public@data)                         #attributes table

#get the centroids of plots
public_centroids <- gCentroid(public, byid=T)
public_centroids@coords
public_centroids@coords[1,]  
public_centroids_line=data.frame(public_centroids@coords, id=1:5)
public_centroids_line<-do.call(rbind, replicate(5, public_centroids_line, simplify=F))

nm=matrix(ncol=3)
for (i in 1:5){
  nm<-rbind(nm,do.call(rbind,replicate(5,as.matrix(public_centroids_line[i,]),simplify=FALSE)))
}
nm<-nm[-1,]

colnames(nm)<-c("x2","y2","id.dest")
newds<-cbind(public_centroids_line,as.data.frame(nm))
newds1<-newds[-which(newds$id==newds$id.dest),]
newds1$id3=abs(newds1$x-newds1$x2)
newds1=newds1[!duplicated(newds1$id3), ]

newds1$x2<-as.numeric(as.character(newds1$x2)) #converting from factor to numeric
newds1$y2<-as.numeric(as.character(newds1$y2))

l <- vector("list", nrow(newds1)) #

newds1$Euclidean=NA
for(i in 1:nrow(newds1)){
  newds1$Euclidean[i]=max(pointDistance(public_centroids[c(newds1[i,3],newds1[i,6])], lonlat=FALSE))
}
newds1=merge(newds1, all.dist, by="Euclidean")

library(sp)
for (i in seq_along(l)) {
  l[[i]] <- Lines(list(Line(rbind(as.matrix(newds1[i,2:3]),as.matrix(newds1[i,5:6])))), as.character(i))
}

l.spatial<-SpatialLines(l) 

#this for question 5 to plot the lines and weights
l.spatial = sp::SpatialLinesDataFrame(l.spatial, data.frame(ID = c(1:10), newds1[,c(1,9:11)]), match.ID = T)

land_cost_gg <- rasterToPoints(land_cost)
land_cost_gg <- as.data.frame(land_cost_gg)
head(land_cost_gg)

range01 <- function(x){(x-min(x))/(max(x)-min(x))}

# plots for each
(euc.plot <- ggplot() +
  geom_raster(data = land_cost_gg, aes(x = x, y = y, fill = as.factor(gfchab_500cl2)))+
  geom_polygon(data = public, aes(x = long, y = lat, group = group), color = "white", fill = "gray") +
  geom_path(data = l.spatial, aes(x = long, y = lat), size = (rep(range01(1/l.spatial$Euclidean), each=2)+.1))+
  coord_equal()+
  theme_classic()+
  labs(y = "Northing (m)", x= "Easting (m)", fill = "Resistance values")+
  scale_fill_viridis_d()+
  theme(legend.position = "none")+
  ggtitle("Euclidean distance")) 

(lcd.plot <- ggplot() +
  geom_raster(data = land_cost_gg, aes(x = x, y = y, fill = as.factor(gfchab_500cl2)))+
  geom_polygon(data = public, aes(x = long, y = lat, group = group), color = "white", fill = "gray") +
  geom_path(data = l.spatial, aes(x = long, y = lat), size = (rep(range01(1/l.spatial$lcd), each=2)+.1))+
  coord_equal()+
  theme_classic()+
  labs(y = "Northing (m)", x= "Easting (m)", fill = "Resistance values")+
  scale_fill_viridis_d()+
  theme(legend.position = "none")+
  ggtitle("Least cost distance"))

(circuit.plot <- ggplot() +
  geom_raster(data = land_cost_gg, aes(x = x, y = y, fill = as.factor(gfchab_500cl2)))+
  geom_polygon(data = public, aes(x = long, y = lat, group = group), color = "white", fill = "gray") +
  geom_path(data = l.spatial, aes(x = long, y = lat), size = (rep(range01(1/l.spatial$circuit), each=2)+.1))+
  coord_equal()+
  theme_classic()+
  labs(y = "Northing (m)", x= "Easting (m)", fill = "Resistance values")+
  scale_fill_viridis_d()+
  theme(legend.position = "none")+
  ggtitle("Circuit theory"))

(rsp.plot <- ggplot() +
  geom_raster(data = land_cost_gg, aes(x = x, y = y, fill = as.factor(gfchab_500cl2)))+
  geom_polygon(data = public, aes(x = long, y = lat, group = group), color = "white", fill = "gray") +
  geom_path(data = l.spatial, aes(x = long, y = lat), size = (rep(range01(1/l.spatial$rSP), each=2)+.1))+
  coord_equal()+
  theme_classic()+
  labs(y = "Northing (m)", x= "Easting (m)", fill = "Resistance values")+
  scale_fill_viridis_d()+
  theme(legend.position = "none")+
  ggtitle("Randomized shortest paths"))
```

## a. Plot
```{r, message = F, message=F, warning=F, fig.dim=c(12, 12)}
egg::ggarrange(euc.plot, lcd.plot, circuit.plot, rsp.plot, nrow = 2)
```

## b. Generally, the patern of which lines are more weighted than others are consistant between all methods. There is a difference where some of the lines are more weighted in the Euclidean panel. 

## c. The conservation area pairing with greatest travel cost by least cost distance is between Big Cypress National Preserve and Kissimmee Prarie Preserve State Park. The conservation area pairing that shows the least travel cost by circuit theory is between Big Cypress National Preserve and Florida Panther National Wildlife Refuge. 


# 6. 
```{r,message = F, message=F, warning=F, fig.show="hide"}

#9.3.3.2 Least-cost paths

#attribute table
pubdata <- public@data

#eyeball where a box is to contain both
# ggplot() +
#   geom_polygon(data = public, aes(x = long, y = lat, group = group), color = "white", fill = "gray")

#crop to focal area: where are the extents that cover both  
fbc_ossf_extent <- extent(600000,680000,275000,335000)

land_sub <- crop(land, fbc_ossf_extent)

land_cost_sub <- crop(land_cost, fbc_ossf_extent)

land_cond_sub <- transition(1/land_cost_sub, transitionFunction=mean, 8)

land_cond_sub <- geoCorrection(land_cond_sub, type="c", multpl=FALSE)

#get lcp
fbc_ossf_lcp <- shortestPath(land_cond, public_centroids@coords[3,], public_centroids@coords[1,], output="SpatialLines")

#plot
plot(land_cost_sub, axes=F, box=F)
plot(public, add=T)
points(public_centroids, col="grey20")
lines(fbc_ossf_lcp, col="red", lw=3)

############################################
#9.3.3.3 Least-cost corridor
############################################
plot(public_centroids)

#get cumulative costs from each PA
fbc.cost <- accCost(land_cond_sub, public_centroids@coords[1,])
ossf.cost <- accCost(land_cond_sub, public_centroids@coords[3,])

#plot
par(mfrow=c(1,2))
plot(fbc.cost)
plot(ossf.cost)
dev.off()

#get least-cost corridor
leastcost_corridor <- overlay(ossf.cost, fbc.cost, fun=function(x, y){return(x + y)})

#plot
plot(leastcost_corridor, legend=F, axes=F)
plot(public, add=T)
points(public_centroids, col="grey30")

#get lower 20% quantile
quantile20 <- quantile(leastcost_corridor, probs=0.20, na.rm=TRUE)

#make new truncated layer
leastcost_corridor20 <- leastcost_corridor

values(leastcost_corridor20) <- NA

leastcost_corridor20[leastcost_corridor < quantile20] <- 1 #truncate to identify corridor

#plot
plot(leastcost_corridor, legend=F, axes=F)
plot(leastcost_corridor20, legend=F,axes=F, add=T)
points(public_centroids, col="grey30")
lines(fbc_ossf_lcp, col="red", lw=3)

############
gg_corridor=as.data.frame(leastcost_corridor, xy=T)
gg_corridor20=as.data.frame(leastcost_corridor20, xy=T)
gg_lcp= sp::SpatialLinesDataFrame(fbc_ossf_lcp, data.frame(ID = c(1)), match.ID = F)

poly_two_park=subset(public, MANAME=="FRED C. BABCOCK-CECIL M. WEBB WILDLIFE MANAGEMENT AREA"|MANAME=="OKALOACOOCHEE SLOUGH STATE FOREST")
```

## Plot
```{r,message = F, message=F, warning=F}
# Plot 
ggplot()+
  geom_raster(data=gg_corridor, aes(x=x, y=y, fill=(layer)))+
  geom_raster(data=na.omit(gg_corridor20), aes(x=x, y=y), fill="gray")+
  geom_path(data=gg_lcp,  aes(x=long, y=lat), size=2, colour="red")+
  geom_point(data=as.data.frame(public_centroids), aes(x=x, y=y),colour="white", size=4)+
  geom_polygon(data=poly_two_park, aes(x=long, y=lat, group=group),colour="white", fill="gray", alpha=.3)+
   coord_equal(xlim=c(min(gg_corridor$x),max(gg_corridor$x)),
              ylim=c(min(gg_corridor$y),max(gg_corridor$y)))+
  viridis::scale_fill_viridis(option = "B", direction=-1)+
  theme_classic()+
  labs(y="Northing (m)", x="Easting (m)", fill="Sum of the \ncumulative \nresistances")+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(text = element_text(size=15))
```

## Relating paths to land-cover types
```{r, message = F, message=F, warning=F}
#identify land-cover along the lcp
lcp.land <- raster::extract(land, fbc_ossf_lcp)

#summarize
table(lcp.land)

#identify land-cover along the least-cost corridor
corridor.land <- mask(land_sub, leastcost_corridor20)

#summarize
table(as.vector(corridor.land))

classification[,1:2]#cross-walk IDs with descriptions
```

## a. The dominant land cover type in the corridor is class 9, or sugarcane based on the descriptions. 
```{r, message = F, message=F, warning=F}
possible_mets <- list_lsm()

# calculate area for each class with lsm
q6a <- lsm_c_ca(corridor.land, directions = 4)

#check
table(values(corridor.land))
```

## b. The Shannon diversity is 2.53 and the evenness is 0.79. 
```{r, message = F, message=F, warning=F}
table(values(corridor.land))

# Diversity is D
C <- table(values(corridor.land))
P <- C / sum(C)
(D <- -sum(P * log(P)))

# Evenness 
(E <- D/log(length(C)))
```


# 7. Flow mapping
```{r, message = F, message=F, warning=F, fig.show="hide"}
# change numbers in brackets to be 1 and 3  
unique(public$DESCRIPT)

#flow mapping under different thetas
passage.map_t0 <- passage(land_cond_sub, origin=public_centroids@coords[1,], goal=public_centroids@coords[3,], theta=0)

passage.map_t000001 <- passage(land_cond_sub, origin=public_centroids@coords[1,], goal=public_centroids@coords[3,], theta=0.000001,totalNet = "total")

passage.map_t00001 <- passage(land_cond_sub, origin=public_centroids@coords[1,], goal=public_centroids@coords[3,], theta=0.00001,totalNet = "total")


passage.map_t001 <- passage(land_cond_sub, origin=public_centroids@coords[1,], goal=public_centroids@coords[3,], theta=0.001)


#plot
plot(passage.map_t0, axes=F, legend=F)
plot(passage.map_t000001, axes=F, legend=F)
plot(passage.map_t00001, axes=F, legend=F)
plot(passage.map_t001, axes=F, legend=F)

#raster to points 
passage.map_t0.pt <- rasterToPoints(passage.map_t0)
passage.map_t0.pt <- as.data.frame(passage.map_t0.pt)

passage.map_t0.pt <- passage.map_t0.pt %>% 
                      rename(theta_0 = layer)
head(passage.map_t0.pt)

#plot
(t_0 <- ggplot()+
  geom_raster(passage.map_t0.pt, mapping = aes(x = x, y = y, fill = theta_0))+
 geom_polygon(data=poly_two_park, aes(x=long, y=lat, group=group),colour="white", fill="gray", alpha=.3)+
  viridis::scale_fill_viridis(option = "B", direction=1)+
  theme_classic()+
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(y="Northing (m)", x="Easting (m)",
       title = expression(Theta*"= 0.00")))
  
# make df
passage.map_t000001.pt <- rasterToPoints(passage.map_t000001)
passage.map_t000001.pt <- as.data.frame(passage.map_t000001.pt)

passage.map_t000001.pt <- passage.map_t000001.pt %>% 
    rename(theta_000001 = layer)

head(passage.map_t000001.pt)

#plot
(t_000001 <- ggplot()+
  geom_raster(passage.map_t000001.pt, mapping = aes(x = x, y = y, fill = theta_000001))+
 geom_polygon(data=poly_two_park, aes(x=long, y=lat, group=group),colour="white", fill="gray", alpha=.3)+
  viridis::scale_fill_viridis(option = "B", direction=1)+
  theme_classic()+
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x="Easting (m)", y= "",
       title = expression(Theta*"= 0.000001")))


#df 
passage.map_t00001.pt <- rasterToPoints(passage.map_t00001)
passage.map_t00001.pt <- as.data.frame(passage.map_t00001.pt)

passage.map_t00001.pt <- passage.map_t00001.pt %>% 
                            rename(theta_00001 = layer)

head(passage.map_t00001.pt)

#plot
(t_00001 <- ggplot()+
  geom_raster(passage.map_t00001.pt, mapping = aes(x = x, y = y, fill = theta_00001))+
 geom_polygon(data=poly_two_park, aes(x=long, y=lat, group=group),colour="white", fill="gray", alpha=.3)+
  viridis::scale_fill_viridis(option = "B", direction=1)+
  theme_classic()+
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x="Easting (m)", y= "",
       title = expression(Theta*"= 0.00001")))

#df 
passage.map_t001.pt <- rasterToPoints(passage.map_t001)
passage.map_t001.pt <- as.data.frame(passage.map_t001.pt)

passage.map_t001.pt <- passage.map_t001.pt %>% 
                            rename(theta_001 = layer)

head(passage.map_t001.pt)

#plot
(t_001 <- ggplot()+
  geom_raster(passage.map_t001.pt, mapping = aes(x = x, y = y, fill = theta_001))+
 geom_polygon(data=poly_two_park, aes(x=long, y=lat, group=group),colour="white", fill="gray", alpha=.3)+
  viridis::scale_fill_viridis(option = "B", direction=1)+
  theme_classic()+
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x="Easting (m)", y= "",
       title = expression(Theta*"= 0.001")))

```

## Plot
```{r, message = F, message=F, warning=F, fig.dim=c(12,12)}

egg::ggarrange(t_0, t_000001, t_00001, t_001, nrow = 1)
```

The theta values signify how the model will converge on a path and how many path options or deviations might be possible. They influence connectivity measures because as theta increases, it forces the connectivity model to converge to a narrower/ straighter least-cost path with fewer path options/ deviations whereas closer to 0 is random. 





