---
title: "ECOL 620 Lab 5"
author: "Natalie Schmer"
date: "2/18/2021"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r, message = F}
#load packages
{
  library(pgirmess)         
  library(ncf)              
  library(spdep)            
  library(geoR)             
  library(gstat)            
  library(RandomFields)     
  library(raster)           
  library(waveslim)         
  library(fields)           
  library(reshape2)         
  library(vegan)            
  library(adespatial)
  library(here)
  library(tidyverse)
  library(RColorBrewer)
}
```


## 1. 
The difference in use between Moran's I and Ripley's K/L functions is that even though both are measures of spatial correlation, Moran's I is more focused on the relationship of the physical locations and characteristics of what is being measured in the locations and the Ripley functions are more focused on the pattern of arrangement of whatever is being measured over different distances. 

## 2. 
In a correlogram, the scale is adjusted to -1:1 and is for understanding autocorrelations and determinig what data to actually use in further analysis. A semivariogram is on a different scale that directly compares variance to distance and scales are only limited by the range of values for distance and variance.  

## 3. 
-a: Y-intercept, above a starts the partial sill and below a is the nugget. 
-b: Sill
-c: Partial sill 
-d: Range 

# Cactus data
## 4. pgirmiss, ncf, spdep corellograms 
```{r, warning=F, message=FALSE, fig.show='hide'}
#load data
matrix <- read.csv('/Users/natalieschmer/Desktop/GitHub/ECOL_620/data/data_for_lab5/cactus_matrix.csv', header=T)

#inspect
head(matrix)

#plot
plot(matrix[,"y"] ~ matrix[,"x"],
     pch=21, cex=1.2,
     bg=gray.colors(12)[cut(matrix[,"Height"], breaks = 12)])

#plot distribution of vegetation height, not normally dist 
hist(matrix[,"Height"], xlab="Vegetation height (cm)",
     main="histogram of vegetation height")

#calculate distance matrix
coords <- cbind(matrix$x, matrix$y)

colnames(coords) <- c("x", "y")
distmat <- as.matrix(dist(coords))

#inspect
dim(distmat)
max(distmat)

#Maximum distance to consider in correlogram/variogram (~1/2 to 2/3 total dist) don't go beyond 47-50 m
maxdist <- 2/3*max(distmat)
maxdist
```

### Calculations
#### Correlogram with pgirmess
```{r, warning=F, message=FALSE, fig.show='hide'}

# using height, moran i, nbclasses is related to pairwise comparisons 
correlog.pgirmess <- pgirmess::correlog(coords, matrix$Height, method="Moran",
                            nbclass=14, alternative = "two.sided")

#Moran and P values for each distance class, coef is moran i, pval- is moran i sig diff than 0, n is number pointwise comparisons in that distance band
round(correlog.pgirmess,2)

#plot: high correlation at 0.19, starts to break down and after 50 m p bad 
plot(correlog.pgirmess[,1], correlog.pgirmess[,2],
     xlab="Distance (m)", ylab="Moran's I", col = ifelse(correlog.pgirmess[,3] < 0.05,'red','black'), pch=19)
abline(h=0)

# make dataframe out of cols 1 and 2
correlog.pgirmess.df <- as.data.frame(correlog.pgirmess) %>% 
                        rename(Distance = dist.class,
                               moran = coef) %>% 
                      mutate(Method = "pgirmess")

```

#### Correlogram with ncf
```{r, warning=F, message=FALSE, fig.show='hide'}
#Correlogram with non-parameteric test of significance, spline and conf int
correlog.ncf <- ncf::correlog(x = matrix$x, y = matrix$y, z = matrix$Height,
                         increment=5, resamp=99)

#plot
plot(correlog.ncf)
abline(h=0)

# make df 
correlog.ncf.df <- as.data.frame(correlog.ncf[c(2,3,5)])

correlog.ncf.df <- correlog.ncf.df %>% 
                      rename(Distance = mean.of.class,
                             Moran = correlation,
                             p.value = p) %>% 
                      mutate(Method = "ncf")
```

#### Correlogram with spline and 95% pointwise bootstrap confidence intervals
```{r, warning=F, message=FALSE, fig.show='hide'}
spline.corr <- spline.correlog(x = matrix$x, y = matrix$y, z = matrix$Height,
                               xmax = maxdist, resamp=99, type="boot")

#plot
plot(spline.corr)

#--------------------------------------#
#Moran's I test with spdep
#--------------------------------------#
# pull out the relevant stuff 
#make a neighborhood list
neigh <- dnearneigh(x=coords, d1=0, d2=3, longlat=F)#d1 is minimum distance, d2 is max distance

#plot the neighorhood
plot(neigh,coordinates(coords))

#create weights for the neighbors
wts <- nb2listw(neighbours=neigh, style='W', zero.policy=T)#W = row-standardized weights

#Moran's I test with normal approximation versus Monte Carlo permutation test
mor.mc <- moran.mc(x=matrix$Height, listw=wts, nsim=999, zero.policy=T) #Monte Carlo
mor.norm <- moran.test(x=matrix$Height, listw=wts, randomisation=F, zero.policy=T)#normal approximation

#inspect
mor.mc
mor.norm

#--------------------------------------#
#Correlograms with spdep
#--------------------------------------#

#repeat above with a for loop using consecutive lag distances

#Create data frame for storing output
correlog.sp <- data.frame(dist=seq(5, maxdist, by=5),
                        Morans.i=NA, Null.lcl=NA, Null.ucl=NA, Pvalue=NA)
#inspect
head(correlog.sp)

#then do a for loop to calculate Moran's I for lag distances
for (i in 1:nrow(correlog.sp)){

  d.start <- correlog.sp[i,"dist"]-5
  d.end <- correlog.sp[i,"dist"]

  neigh <- dnearneigh(x=coords, d1=d.start, d.end, longlat=F)
  wts <- nb2listw(neighbours=neigh, style='W', zero.policy=T)
  mor.i <- moran.mc(x=matrix$Height, listw=wts, nsim=99, alternative="greater", zero.policy=T)

  #summarize results from spdep
  correlog.sp[i, "dist"] <- (d.end+d.start)/2                                    #mean dist
  correlog.sp[i, "Morans.i"] <- mor.i$statistic 								                 #observed Moran's I
  correlog.sp[i, "Null.lcl"] <- quantile(mor.i$res, probs = 0.025,na.rm = TRUE)  #lower null envelope
  correlog.sp[i, "Null.ucl"] <- quantile(mor.i$res, probs = 0.975,na.rm = TRUE)  #upper null envelope
  correlog.sp[i, "Pvalue"] <- mor.i$p.value									                     #p-value for Moran's I at that distance category
}

#plot
plot(y=correlog.sp$Morans.i, x=correlog.sp$dist,
     xlab="Lag Distance(m)", ylab="Moran's I", ylim=c(-0.3,0.3))         #ylim provides limit on y-axis between -1 and 1
abline(h=0)                                                              #0 reference
lines(correlog.sp$dist, correlog.sp$Null.lcl,col = "red")	               #add the null lcl to the plot
lines(correlog.sp$dist, correlog.sp$Null.ucl,col = "red")	               #add the null ucl to the plot

# add identifier to correlog.sp
correlog.sp.df <- correlog.sp %>% 
                  mutate(Method = "spdep") %>% 
                  rename(p.value = Pvalue,
                         Distance = dist,
                         Moran = Morans.i)

```


### Format each df to rbind to plot  
```{r, warning=F, message=FALSE}
# check names and fix and select cols 
sort(names(correlog.pgirmess.df))
sort(names(correlog.ncf.df))
sort(names(correlog.sp.df))

correlog.pgirmess.df <- correlog.pgirmess.df %>% 
                          dplyr::select(Distance, Method, Moran = moran, p.value)

correlog.ncf.df <- correlog.ncf.df %>% 
                          dplyr::select(Distance, Method, Moran, p.value)

correlog.sp.df <- correlog.sp.df %>% 
                    dplyr::select(Distance, Method, Moran, p.value)

# ribind to plot out of this df 
all.corrs.df <- rbind(correlog.pgirmess.df, correlog.ncf.df, correlog.sp.df )
```

## Plot 
```{r, warning=F, message=FALSE}
all.corrs.df %>% 
  dplyr::filter(Distance <= maxdist) %>% 
  ggplot(., mapping = aes(Distance, Moran, shape = Method)) +
    geom_line() +
    geom_point(. %>% filter(p.value < 0.05), mapping =aes(),color = "red", size = 3) +
    geom_point(. %>% filter(p.value > 0.05), mapping = aes(), color = "black", size = 3) +
    geom_hline(yintercept = 0) +
    theme_bw()+
    labs(x = "Distance (m)",
         y = "Moran's I")
```

Spatial autocorrelation happens for all methods up to D = ~15 m, and then after that is more sporadic and depends on the method with no similar distances having all methods showing autocorrelation. 

# Cactus data: Semivariogram 
## 5. gstat
```{r, warning=F, message=FALSE}
gstat.veg <- matrix
coordinates(gstat.veg) = ~x + y

#Empirical semivariogram
emp.gstat <- variogram(Height ~ 1, cutoff=maxdist, width=3, gstat.veg)

#plot variogram
plot(emp.gstat)
```

Based on this plot, I would say that the nugget is from y = 0 to y = ~920, the sill is around y = 1260, and the range is a little less than 30 m, maybe around 28. 

## 6. Anisotropy 
```{r, warning=F, message=FALSE, fig.show='hide'}
#create geoR object
geoR.veg <- as.geodata(matrix)

#plot
plot(geoR.veg)

#Empirical semivariogram
emp <- variog(geoR.veg, max.dist=maxdist)

#plot, y is semivariance, plateau 
plot(emp)

#standardize breaks
emp <- variog(geoR.veg, max.dist=maxdist, breaks=c(seq(0,maxdist,by=3)))

#plot variogram
(plot(emp))
```

```{r, warning=F, message=FALSE}
#Directional variogram in geoR
emp4 <- variog4(geoR.veg, max.dist=maxdist)

#plot directional variogram
(plot(emp4))

#Directional variogram in gstat
emp4.gstat <- variogram(Height ~ 1, cutoff=maxdist, alpha=c(0,45,90,135), gstat.veg)

#plot directional variogram
(plot(emp4.gstat))
```

Based on these plots, I think there is some concern about anisotropy because the lines all seem to deviate from each other pretty quickly after about distance = 10 m and it is hard to discern a pattern that applies to all lines. 

## 7. Variogram plot 
```{r, warning=F, message=FALSE, fig.show='hide'}
#Model-based variograms with likelihood fitting in geoR

#Exponential variogram
mlexp <- likfit(geoR.veg, cov.model="exp", ini=c(700,10))

#Spherical variogram
mlsph <- likfit(geoR.veg, cov.model="sph", ini=c(700,10))

#inspect
summary(mlexp)
summary(mlsph)

#model selection tool to look for models w lower aic values
AIC(mlexp,mlsph)  

#plot
plot(emp)
lines(mlexp, col="blue")
lines(mlsph, col="red")

#Monte Carlo envelopes
emp.env <- variog.mc.env(geoR.veg, obj.var=emp)

#plot
plot(emp, envelope=emp.env)
lines(mlsph, col="red")
lines(mlexp, col="blue")


#Model-based variograms with least squares in gstat

#inspect models that gstat can fit
vgm()#fits more models than geoR
show.vgms()#plots examples of the various variograms

#Spherical variogram
sph.gstat <- fit.variogram(emp.gstat, vgm("Sph")) #in vgm(psill, model, range, nugget)

#Exponential variogram
exp.gstat <- fit.variogram(emp.gstat, vgm("Exp"))

#inspect
exp.gstat
sph.gstat

#plot
plot(emp.gstat, exp.gstat)
plot(emp.gstat, sph.gstat)
```

### Extract from sph.gstat and plot
```{r, warning=F, message=FALSE}
?variogramLine

#preds[1,2]+sph.gstat$psill[2] 
#Extract theoretical 
sph.theo <- variogramLine(sph.gstat, maxdist = maxdist, min = 1.0e-6 * maxdist, 
	dir = c(1,0,0), covariance = FALSE)

#get value for where red line should be, distance where the plateau/ sill starts 
mindist <- sph.theo %>% 
  filter(gamma == max(gamma)) %>% 
  filter(dist == min(dist)) %>% 
  select(dist) %>% 
  as.numeric()

sph.gstat$range
#red line/ max is 15.997

sph.gstat$psill
#blue line/ nugget is at 820.5813
#green line/ sill is at 1287.013 aka max gamma 

ggplot()+
geom_point(data = emp.gstat, aes(x=dist, y=gamma, size = np))+
  geom_hline(yintercept = 820.5813, color = "blue")+
  geom_hline(yintercept = max(sph.theo$gamma), color = "green")+
  geom_line(sph.theo, mapping = aes(dist, gamma))+
  geom_vline(xintercept = 15.997, color = "red")+
  labs(x= "Distance (m)",
       y = "Semivariance")+
  theme_bw()
```

## 8. Inverse distance and Kriging 
```{r, warning=F, message=FALSE}
#Create grid with intervals of 0.5 unit (.5-m)
#need labels for coords
new.grid.5m <- expand.grid(x=seq(0,max(matrix$x),.5), 
                           y=seq(0,max(matrix$y),.5))

#Maximum likelihood estimates from the variogram modeling
mlexp$nugget # nugget
mlexp$cov.pars[1] #partial sill
mlexp$cov.pars[2] #range

#Ordinary kriging w geoR
krig.geoR.exp <- krige.conv(geoR.veg, locations=new.grid.5m, 
                            krige=krige.control(cov.pars=c(mlexp$cov.pars[1], mlexp$cov.pars[2]), 
                                                nugget = mlexp$nugget,
                                              cov.model="exp", type.krige="OK"))

#inspect
hist(krig.geoR.exp$predict)
hist(matrix$Height)

#plot
#see where points were collected 
image(krig.geoR.exp, main="krigged estimates") 
image(krig.geoR.exp, val=sqrt(krig.geoR.exp$krige.var), main="kriging SE")

#kriging with gstat, need labels for coords
new.grid.1m.gstat <- expand.grid(x=seq(0,max(matrix$x),.5), y=seq(0,max(matrix$y),.5))

#convert to sp object
gridded(new.grid.1m.gstat) = ~x + y

# Ordinary kriging w gstat
krig.gstat <- krige(Height ~ 1, 
                    gstat.veg, 
                    new.grid.1m.gstat, 
                    model = sph.gstat)

#Plot
image(krig.gstat, main="krigged estimates-gstat")

image(krig.geoR.exp, main="krigged estimates-geoR")

cor(krig.geoR.exp$predict, krig.gstat$var1.pred)

#Inverse distance weighting in gstat
#idp-the power used; default power=2
idw.gstat <- idw(Height ~ 1, gstat.veg, new.grid.1m.gstat)

image(idw.gstat, main="idw estimates-gstat")
```

Based on the correlations and the plots, the ordinary kriging methods using exponential fit create the best interpolation. 

```{r, warning=F, message=FALSE}
#Correlation between idw, kriging in gstat, kriging in geoR
round(cor(cbind(geoR.exp = krig.geoR.exp$predict, 
                gstat.exp = krig.gstat$var1.pred,
                gstat.idw = idw.gstat$var1.pred)),3)
```


# Elevation data
## 9.  
```{r, warning=F, message=FALSE, fig.show='hide'}
#download elevation code
#change the directory as needed
srtm <- raster("~/Desktop/GitHub/ECOL_620/data/data_for_lab5/srtm_15_05.tif")
e <- extent(-107.5, -107, 37.5, 38)
srtm=crop(srtm, e)
plot(srtm)

#aggregate to reduce computation
srtm = aggregate(srtm, fact=10, fun=mean) #mean value
plot(srtm)

#extract 500 points of elevation, randomly drawn
set.seed(2020)

#xy = T means keep x and y coords
srtm_pts <- as.data.frame(sampleRandom(srtm, 500, xy=T)) 

srtm_df <- rasterToPoints(srtm)
srtm_df <- as.data.frame(srtm_df)

colnames(srtm_pts)[3] <- "elevation"
head(srtm_pts)
plot(srtm)
points(srtm_pts$x,srtm_pts$y)

#Create grid with intervals of 0.008333333 degree, predict kriging back onto this new grid 

extent(srtm)
res(srtm)
e <- extent(srtm)
new.grid<- raster(e, resolution=res(srtm))
new.grid <- as.data.frame(new.grid, xy=TRUE)
new.grid <- new.grid[,1:2]

plot(new.grid)

# make the srtm_pts a matrix
srtm_pts_mat <- as.matrix(srtm_pts)
head(srtm_pts_mat)

plot(srtm_pts_mat[,"y"] ~ srtm_pts_mat[,"x"],
pch =21, bg = gray.colors(12)[cut(srtm_pts_mat[,3], breaks = 12)])

# use 2/3 total distance 
coords_9 <- cbind(srtm_pts$x, srtm_pts$y) 
colnames(coords_9) <- c("x", "y")
distmat_9 <- as.matrix(dist(coords_9))

#maximum distance to consider in variogram
maxdist_9 <- 2/3 * max(distmat)
```

### a: Variogram models 
```{r, warning=F, message=FALSE}
gstat.veg_9 <- srtm_pts

coordinates(gstat.veg_9) <- ~x + y

# change the width
emp.gstat_9 <- variogram(elevation ~ 1, 
                         cutoff = maxdist_9, 
                         width = .05, 
                         gstat.veg_9)

plot(emp.gstat_9)

show.vgms()
```

Visually, the models that seem like the best fit for our variogram are exp, sph, gau, and wav. However, I could not get gau or wav to actually work so I only tested exp and sph. 

```{r}
geor_9 <- as.geodata(gstat.veg_9)

#Model-based variograms with likelihood fitting in geoR

#Exponential variogram
mlexp_9 <- likfit(geor_9, cov.model="exp", ini=c(700,10))

#Spherical variogram
mlsph_9 <- likfit(geor_9, cov.model="sph", ini=c(700,10))

#inspect
summary(mlexp_9)
summary(mlsph_9)

(AIC(mlexp_9, mlsph_9)) 

```

### b. Kriging with exp

I chose exp because had the lower AIC value. 
```{r, warning=F, message=FALSE}
krig.geoR.exp_9 <- krige.conv(geor_9, locations = new.grid,
krige = krige.control(cov.pars = c(mlexp_9$cov.pars[1], mlexp_9$cov.pars[2]), nugget = mlexp_9$nugget,
cov.model = "exp", type.krige = "OK"))

predicted_9 <- krig.geoR.exp_9$predict
original <- srtm_df$srtm_15_05

q9_elev_krig <- cbind(predicted_9, original)

q9_elev_krig <- q9_elev_krig  %>% 
              as.data.frame() %>% 
              mutate(diff = round(predicted_9 - original, 2))

# figure out which ones were sampled, should be 500 bc we sampled 500 
sampled_k <- q9_elev_krig %>% 
              filter(diff == 0 | diff == 0.00)

# remove to correlate 
q9_elev_nosample_k <- q9_elev_krig %>% 
                      filter(diff != 0 | diff != 0.00)

# correlate 
(krig_corr <- cor(q9_elev_nosample_k$predicted_9, q9_elev_nosample_k$original))

#create df to plot from
krig_raster <- cbind(srtm_df, predicted_9)

```

The correlation between the best kriged model prediction and the known elevation is `r toString(krig_corr)`. 

### c. Inverse distance weighting
```{r, warning=F, message=FALSE}
gridded(new.grid) <- ~x+y

idw.gstat_9 <- idw(elevation ~ 1, gstat.veg_9, new.grid)

#extract idw predictions 
idw_pred_9 <- idw.gstat_9$var1.pred

idw_pred_og_9 <- cbind(idw_pred_9, original)

idw_pred_og_9 <- idw_pred_og_9 %>% 
                    as.data.frame() %>% 
                    mutate(diff = round(idw_pred_9 - original,2))

# remove to correlate 
q9_elev_nosample_idw <- idw_pred_og_9 %>% 
                      filter(diff != 0 | diff != 0.00)

# correlate 
(idw_corr <- cor(q9_elev_nosample_idw$idw_pred_9, q9_elev_nosample_idw$original))

#create df to plot from
idw_raster <- cbind(srtm_df, idw_pred_9)

```

My kriging model does perform better than inverse distance weighting: the correlation between the best kriged model prediction and the known elevation is `r toString(krig_corr)` whereas idw is `r toString(idw_corr)`. 

### d. Plots
```{r, warning=F, message=FALSE, fig.show=F}
# create color scale 
# r color brewer spectral
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

#fill
sf <- scale_fill_gradientn(colours = myPalette(100), limits = c(2450.91, 4145.64))

#points 
sp <- scale_colour_gradientn(colours = myPalette(100), limits = c(2450.91, 4145.64))

# Plot of sample points 
sample_points_plot <- ggplot(srtm_pts, aes(x, y, color = elevation))+
  geom_point()+
  labs(fill = "Elevation (m)",
       title = "Sampling Points",
       x = "Longitude (°)",
       y = "Latitude (°)")+
  sp

# Plot of Kriging
krig_plot <- ggplot(srtm_df, aes(x, y, fill = predicted_9))+
  geom_tile()+
  labs(fill = "Elevation (m)",
       title = "Kriging",
       x = "Longitude (°)",
       y = "Latitude (°)")+
  sf

#Plot of IDW
idw_plot <- ggplot(idw_raster, aes(x, y, fill = idw_pred_9))+
  geom_tile()+
  labs(fill = "Elevation (m)",
       title = "Inverse Distance Weighting",
       x = "Longitude (°)",
       y = "Latitude (°)")+
  sf


# Plot of true elevation 
true_elev_plot <- ggplot(srtm_df, aes(x, y, fill = srtm_15_05))+
  geom_tile()+
  labs(fill = "Elevation (m)",
       title = "True Elevation",
       x = "Longitude (°)",
       y = "Latitude (°)")+
  sf
```

```{r, warning=F, message=FALSE, fig.dim=c(12,12)}
cowplot::plot_grid(sample_points_plot, krig_plot, idw_plot, true_elev_plot, nrow = 2, ncol = 2)
```

